---
title: "XAI3"
author: "María Dolores Bonet, Claudia Piqueras and Elena Moya"
date: "2025-05-15"
output: html_document
---


**XAI: Interpretable models 3**

**Load libraries**

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,   
  message = FALSE,  
  warning = FALSE   
)
```

```{r}
library(randomForest)
library(pdp)
library(ggplot2)
library(viridis)
library(ggExtra)
```

**1. 1D PDP for bike rental counts**

*1.1. Read data*

```{r}
day_data <- read.csv("day.csv", stringsAsFactors = FALSE)
```

*1.2. Create “days since 2011” variable*

```{r}
day_data$days_since_2011 <- as.numeric(
  as.Date(day_data$dteday) - as.Date("2011-01-01")
)
```

*1.3. Fit Random Forest*

```{r}
set.seed(123)
rf_bike <- randomForest(
  cnt ~ days_since_2011 + temp + hum + windspeed,
  data = day_data,
  importance = TRUE
)
```

*1.4. Compute 1D partial dependences*

```{r}
p_days <- partial(rf_bike, pred.var = "days_since_2011", grid.resolution = 100)
p_temp <- partial(rf_bike, pred.var = "temp",              grid.resolution = 100)
p_hum  <- partial(rf_bike, pred.var = "hum",               grid.resolution = 100)
p_wind <- partial(rf_bike, pred.var = "windspeed",         grid.resolution = 100)
```


*1.5. Plot them side by side*

```{r}
p1 <- autoplot(p_days, 
               xlab = "Days since 2011", 
               ylab = "Pred. cnt") +
  ggtitle("Days") +
  geom_rug(data = day_data, 
           aes(x = days_since_2011), 
           inherit.aes = FALSE,
           sides = "b",    # bottom rug
           alpha = 0.3)

# Temperature
p2 <- autoplot(p_temp, 
               xlab = "Temperature (norm.)", 
               ylab = "Pred. cnt") +
  ggtitle("Temp") +
  geom_rug(data = day_data, 
           aes(x = temp), 
           inherit.aes = FALSE,
           sides = "b",
           alpha = 0.3)

# Humidity
p3 <- autoplot(p_hum, 
               xlab = "Humidity (norm.)", 
               ylab = "Pred. cnt") +
  ggtitle("Hum") +
  geom_rug(data = day_data, 
           aes(x = hum), 
           inherit.aes = FALSE,
           sides = "b",
           alpha = 0.3)

# Windspeed
p4 <- autoplot(p_wind, 
               xlab = "Windspeed (norm.)", 
               ylab = "Pred. cnt") +
  ggtitle("Wind") +
  geom_rug(data = day_data, 
           aes(x = windspeed), 
           inherit.aes = FALSE,
           sides = "b",
           alpha = 0.3)

# Arrange side by side
gridExtra::grid.arrange(p1, p2, p3, p4, nrow = 2)
```



**2. 2D PDP: Humidity vs Temperature**

*2.1. Fit Random Forest*

```{r}
set.seed(123)
rf_bike <- randomForest(
  cnt ~ days_since_2011 + temp + hum + windspeed,
  data       = day_data,
  importance = TRUE
)
```

*2.2. Sample safely (≤ 2000 rows)*

```{r}
n           <- nrow(day_data)
target_size <- 2000
sample_size <- min(target_size, n)
set.seed(42)
sample_data <- day_data[sample(seq_len(n), sample_size), ]
```

*2.3. Compute 2D PDP*

```{r}
p2d <- partial(
  object          = rf_bike,
  pred.var        = c("temp", "hum"),
  grid.resolution = 50,
  train           = sample_data
)
```

*2.4. Make heatmap + rug marginal*

```{r}
ggplot(p2d, aes(x = temp, y = hum, fill = yhat)) +
  geom_tile(
    width  = diff(range(p2d$temp)) / (50 - 1),
    height = diff(range(p2d$hum))  / (50 - 1)
  ) +
  # Rug marginal: marcas en ambos ejes para cada observación
  geom_rug(
    data         = sample_data,
    aes(x = temp, y = hum),
    inherit.aes  = FALSE,
    alpha        = 0.3,
    size         = 0.5
  ) +
  scale_fill_viridis(
    name   = "Pred. count",
    option = "B"
  ) +
  labs(
    x     = "Temp (normalized)",
    y     = "Hum (normalized)",
    title = "2D PDP of Predicted Bike Rentals\nwith Marginal Rugs"
  ) +
  theme_minimal()
```

**3. One‐dimensional PDP for House Price Prediction**

*3.1. Read data*

```{r}
house_data <- read.csv("kc_house_data.csv", stringsAsFactors = FALSE)
```

*3.2. Safe subsample (≤2 000 rows)*

```{r}
set.seed(42)
n           <- nrow(house_data)
sample_size <- min(2000, n)
sample_data <- house_data[sample(seq_len(n), sample_size), ]
```

*3.3. Fit Random Forest*

```{r}
set.seed(123)
rf_house <- randomForest(
  price ~ bedrooms + bathrooms + sqft_living + sqft_lot + floors + yr_built,
  data       = sample_data,
  importance = TRUE
)
```

*3.4. Compute 1-D partial dependences*

```{r}
p_bedrooms    <- partial(rf_house,
                         pred.var        = "bedrooms",
                         grid.resolution = length(unique(sample_data$bedrooms)),
                         train           = sample_data)
p_bathrooms   <- partial(rf_house,
                         pred.var        = "bathrooms",
                         grid.resolution = 50,
                         train           = sample_data)
p_sqft_living <- partial(rf_house,
                         pred.var        = "sqft_living",
                         grid.resolution = 100,
                         train           = sample_data)
p_floors      <- partial(rf_house,
                         pred.var        = "floors",
                         grid.resolution = length(unique(sample_data$floors)),
                         train           = sample_data)
```

*3.5. Plot each PDP with a rug of the data distribution*

```{r}
p1 <- autoplot(p_bedrooms) +
  ggtitle("Bedrooms") +
  xlab("Number of bedrooms") +
  ylab("Predicted price") +
  geom_rug(data         = sample_data,
           aes(x = bedrooms),
           inherit.aes  = FALSE,
           sides        = "b",
           alpha        = 0.3)

p2 <- autoplot(p_bathrooms) +
  ggtitle("Bathrooms") +
  xlab("Number of bathrooms") +
  ylab("Predicted price") +
  geom_rug(data         = sample_data,
           aes(x = bathrooms),
           inherit.aes  = FALSE,
           sides        = "b",
           alpha        = 0.3)

p3 <- autoplot(p_sqft_living) +
  ggtitle("Sqft living") +
  xlab("Square feet of living area") +
  ylab("Predicted price") +
  geom_rug(data         = sample_data,
           aes(x = sqft_living),
           inherit.aes  = FALSE,
           sides        = "b",
           alpha        = 0.3)

p4 <- autoplot(p_floors) +
  ggtitle("Floors") +
  xlab("Number of floors") +
  ylab("Predicted price") +
  geom_rug(data         = sample_data,
           aes(x = floors),
           inherit.aes  = FALSE,
           sides        = "b",
           alpha        = 0.3)
```

*3.7. Arrange plots in a 2×2 grid*

```{r}
gridExtra::grid.arrange(p1, p2, p3, p4, nrow = 2)
```













